---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
# set time zone to GMT and loading package
Sys.setenv(TZ='GMT')
library(tidyverse)
library(hms)
```

## Loading and preprocessing the data

Downloading, reading, processing data

```{r}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
mydata <- read_csv(unzip(temp, "activity.csv"), na = c("NA",NA), col_names = TRUE  )
unlink(temp)
mydata <- tbl_df(mydata)
head(mydata)
```



## What is mean total number of steps taken per day?


Histogram of total number of steps taken each day.

```{r}
steps_per_day = aggregate(steps ~ date, data = mydata, sum, na.rm = TRUE)
ggplot(data = steps_per_day, aes(x= steps)) + geom_histogram(fill="red", colour = "white") + theme_bw()
```


Mean and median number of steps for each day

```{r}
summary(steps_per_day$steps )
```



## What is the average daily activity pattern?

Time series plot

```{r}
avg_per_day <-aggregate(steps ~ date, data = mydata, mean, na.rm = TRUE)
with(data = avg_per_day, 
        plot( steps, type='l', 
             xlab='Date',
             lwd=2, 
             ylab='Average number of steps',
             main ='Average number of steps taken each day', col=rgb(0.2,0.4,0.6,0.8)
     ))
```


The 5-minute interval that, on average, contains the maximum number of steps:

```{r}
avg_per_interval <-aggregate(steps ~ interval, data = mydata, mean, na.rm = TRUE)
which.max(avg_per_interval$steps)
as_hms(104*5*60)
```

The 5-minute interval that, on average, contains the maximum number of steps is the 104 i.e between 8:35 and 8:40

## Imputing missing values

The total number of missing values in the dataset

```{r}
NA_row <- sum(is.na(mydata))
NA_row
```
The strategy is to fill in a missing value with the average number of steps in the same 5-min interval.
We will create a new dataset as the original and use tapply for the filling in.

```{r}
mydata02 <- mydata
na_row <- is.na(mydata02$steps)
avg_interval <- tapply(mydata02$steps, mydata02$interval, mean, na.rm=TRUE, simplify=TRUE)
mydata02$steps[na_row] <- avg_interval[as.character(mydata02$interval[na_row])]
```


Verifying the result:
```{r}
sum(is.na(mydata))
sum(is.na(mydata02))
```

Histogram of the filled dataset (without NA)

```{r}
head(mydata02)
steps_per_day02 = aggregate(steps ~ date, data = mydata02, sum, na.rm = TRUE)
ggplot(data = steps_per_day02, aes(x= steps)) + geom_histogram(fill="red", colour = "white") + theme_bw()
```


Mean and median number of steps taken each day

```{r}
summary(steps_per_day02$steps )
```

Time series plot:
```{r}
avg_per_interval02 <-aggregate(steps ~ interval, data = mydata02, sum, na.rm = TRUE)
with(data = avg_per_interval02 , 
     plot( steps, col="red", type='l', 
          xlab='Intervals',
          lwd=2, 
          ylab='Average number of steps',
          main ='Average number of steps taken in 5-minute interval'
           ))
```

## Are there differences in activity patterns between weekdays and weekends?


To determine this:
```{r}
mydata02 <- mutate(mydata02, type_of_day = ifelse(weekdays(mydata02$date) == "Saturday" | weekdays(mydata02$date) == "Sunday", "weekend", "weekday"))
mydata02$type_of_day <- as.factor(mydata02$type_of_day)
head(mydata02)
weekday_act <- filter(mydata02,type_of_day == "weekday" )
weekend_act <- filter(mydata02,type_of_day == "weekend" )
avg_per_weekday <-aggregate(steps ~ interval, data = weekday_act, mean, na.rm = TRUE)
avg_per_weekend <-aggregate(steps ~ interval, data = weekend_act, mean, na.rm = TRUE)
par(mfrow = c(2,1))
with(data = avg_per_weekday, plot(steps, type='l', 
          xlab='Interval (Weekday)',
          lwd=2, 
          ylab='Average number of steps',
          main ='Average number of steps taken in weekdays',
          col=rgb(0.4,0.9,0.4,0.8)))
with(data = avg_per_weekend, plot(steps, type='l',
          xlab='Interval (Weekend)',
          lwd=2, 
          ylab='Average number of steps',
          main ='Average number of steps taken in weekend', 
          col=rgb(0.9,0.4,0.4,0.8)))
```

The graphs display clear differences in patterns between

rmarkdown::render("PA1_template.Rmd")

